---
title: "CoMorTrax: Comorbidity-Aware Multi-Scale Transcriptomic Analysis"
author:
  - name: Ahmed Mokhtar
    affiliation: Bioinformatics Research Group
    email: ahmed.salem@bioinformatics.org
date: "`r Sys.Date()`"
package: CoMorTrax
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{CoMorTrax: Comorbidity-Aware Multi-Scale Transcriptomic Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.align = "center",
  warning  = FALSE,
  message  = FALSE,
  eval     = FALSE
)
```

# Introduction

## The Problem of Comorbidity in Single-Cell Genomics

Comorbidities -- the co-occurrence of two or more diseases in the same patient --
affect the majority of adults over 65 and are responsible for disproportionately
poor clinical outcomes. Despite this, almost all single-cell computational tools
analyze diseases in isolation: "Disease A vs. Control" or "Disease B vs. Control."

This reductionist approach misses the point. In a patient with both
Alzheimer's disease (AD) and Type 2 Diabetes (T2D), the transcriptomic signature
of neurons or microglia is **not** simply the sum of the AD and T2D signatures.
Regulatory networks rewire. Pathways are synergistically amplified or antagonistically
suppressed. These non-additive interactions are the molecular basis of why
comorbid patients have worse outcomes than predicted by either condition alone.

## What CoMorTrax Does

**CoMorTrax** (Comorbidity Tracking) is a Bioconductor framework that explicitly
models these multi-disease molecular interactions at single-cell resolution.

### Workflow Diagram

```
Raw scRNA-seq + Multi-Disease Metadata
              |
              v
  +---------------------------+
  |  1. Object Construction   |  createCoMorTraxObject()
  +---------------------------+
              |
              v
  +---------------------------+
  |  2. Preprocessing         |  preprocessCoMorTrax()
  |  QC, Norm, HVG, PCA, UMAP|
  +---------------------------+
              |
              v
  +---------------------------+
  |  3. Disease Encoding      |  encodeDiseaseLabels()
  |  Binary multi-label matrix|
  +---------------------------+
              |
              v
  +---------------------------+
  |  4. Comorbidity DE        |  runComorbidityDE()
  |  All comparisons, by type |
  +---------------------------+
              |
              v
  +============================+
  |  5. INTERACTION EFFECTS   |  quantifyInteractionEffects()  [CORE]
  |  Delta = obs - expected   |
  |  Synergistic / Antagonist |
  +============================+
         /          \
        v            v
  +----------+  +------------------+
  | Pathways |  | Cell Vulnerability|
  | Module 4 |  | Module 5          |
  +----------+  +------------------+
        \            /
         v          v
  +---------------------------+
  |  6. Network Rewiring      |  buildComorbidityNetworks()
  |  GRN comparison per state |
  +---------------------------+
              |
              v
  +---------------------------+
  |  7. ML Classification     |  trainComorbidityClassifier()
  |  RF / LASSO / ElasticNet  |
  +---------------------------+
              |
              v
  +---------------------------+
  |  8. Export + Validation   |  exportCoMorTrax()
  +---------------------------+
```

## Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("CoMorTrax")
```

For the development version:

```{r install-dev, eval=FALSE}
BiocManager::install("AhmedSalem/CoMorTrax")
```

## Loading the Package

```{r load}
library(CoMorTrax)
library(SingleCellExperiment)
library(S4Vectors)
library(SummarizedExperiment)
```

---

# Simulating Example Data

CoMorTrax ships with a simulation engine that generates realistic comorbidity
data with **injected** synergistic and antagonistic signals of known identity,
enabling truth-recoverable benchmarking.

```{r simulate}
set.seed(42)
sce <- simulateCoMorTrax(
  n_cells    = 400,
  n_genes    = 300,
  n_syn      = 20,    # genes synergistically amplified under comorbidity
  n_ant      = 15,    # genes antagonistically suppressed under comorbidity
  diseases   = c("AD", "T2D"),
  cell_types = c("Neuron", "Microglia", "Astrocyte")
)

# True interaction gene identities are stored in metadata
true_syn <- metadata(sce)$sim_params$syn_genes
true_ant <- metadata(sce)$sim_params$ant_genes
cat("True synergistic genes:", paste(head(true_syn, 5), collapse = ", "), "\n")
cat("True antagonistic genes:", paste(head(true_ant, 5), collapse = ", "), "\n")

sce
```

---

# Full Pipeline

## Step 1: Create CoMorTraxObject

```{r create}
cmt <- createCoMorTraxObject(sce)
cmt
```

## Step 2: Preprocessing

```{r preprocess}
cmt <- preprocessCoMorTrax(
  cmt,
  normMethod   = "logNorm",
  nHVG         = 2000,
  nPCs         = 30,
  runUMAP      = TRUE,
  batchCorrect = FALSE,
  verbose      = TRUE
)
```

## Step 3: Disease Encoding

Encode each cell's disease state into a binary matrix. Cells with column
`disease_AD = 1` belong to the AD group; comorbid cells have both set to 1.

```{r encode}
cmt <- encodeDiseaseLabels(
  cmt,
  diseaseCol   = c("disease_AD", "disease_T2D"),
  controlLabel = "control",
  verbose      = TRUE
)

# Inspect the binary disease matrix
head(diseaseMatrix(cmt))

# All automatically enumerated comparisons
listComorbidityComparisons(cmt)
```

## Step 4: Comorbidity-Aware Differential Expression

Run DE analysis across **all** comparisons simultaneously, stratified by
cell type.

```{r de}
cmt <- runComorbidityDE(
  cmt,
  method      = "wilcoxon",
  cellTypeCol = "cellType",
  verbose     = TRUE
)

# Access results for a specific comparison
de <- getDeTable(cmt)
head(de[de$significant, ])
```

## Step 5: Interaction Effect Quantification

This is the analytical core of CoMorTrax. For each gene, compute:

$$\Delta = FC(D_1 \cap D_2\,\text{vs. ctrl}) - [FC(D_1\,\text{vs. ctrl}) + FC(D_2\,\text{vs. ctrl})]$$

Genes with Delta > threshold are **synergistic**; those below are
**antagonistic**; the rest are **additive**.

```{r interaction}
cmt <- quantifyInteractionEffects(
  cmt,
  nPermutations = 1000,
  synThreshold  = 0.25,
  antThreshold  = -0.25,
  verbose       = TRUE
)

# Access results
int_df <- interactionScores(cmt)
table(int_df$classification)

# Retrieve gene lists
syn_genes <- getSynergisticGenes(cmt)
ant_genes <- getAntagonisticGenes(cmt)
cat("Synergistic genes detected:", length(syn_genes), "\n")

# Benchmark: compare detected vs true injected
true_syn <- metadata(sce)$sim_params$syn_genes
cat("Precision:", round(mean(syn_genes %in% true_syn), 3), "\n")
cat("Recall:",    round(mean(true_syn %in% syn_genes), 3), "\n")
```

## Step 6: Pathway-Level Comorbidity Enrichment

```{r pathways}
cmt <- runComorbidityPathways(
  cmt,
  genesets = c("GO_BP", "KEGG"),
  organism = "hsa",
  pCutoff  = 0.05,
  verbose  = TRUE
)

# Compare enriched pathways between synergistic and antagonistic sets
comparePathways(cmt, topN = 10)
```

## Step 7: Cell-Type Vulnerability Scoring

```{r vulnerability}
cmt <- scoreCellVulnerability(
  cmt,
  cellTypeCol         = "cellType",
  includePathwayScore = TRUE,
  weights             = list(synergistic  = 0.35,
                              antagonistic = 0.25,
                              pathway      = 0.25,
                              magnitude    = 0.15),
  verbose             = TRUE
)

cellVulnerability(cmt)
```

## Step 8: Gene Regulatory Network Modeling

```{r networks}
cmt <- buildComorbidityNetworks(
  cmt,
  method          = "correlation",
  corrThreshold   = 0.6,
  computeRewiring = TRUE,
  verbose         = TRUE
)

# Retrieve hub genes for the comorbid state
hubs <- getNetworkHubs(cmt, state = "AD_T2D", topN = 10)
print(hubs)

# Network rewiring between control and comorbid state
rew <- computeRewiring(cmt, state1 = "control", state2 = "AD_T2D")
cat("Edge Jaccard (control vs AD+T2D):", round(rew$jaccardEdges, 3), "\n")
cat("Gained edges:", rew$gainedEdges, "| Lost edges:", rew$lostEdges, "\n")
```

## Step 9: Machine Learning Classification

```{r ml}
cmt <- trainComorbidityClassifier(
  cmt,
  method      = "randomForest",
  featureType = "interaction",
  cvFolds     = 5L,
  verbose     = TRUE
)

clf <- classifier(cmt)
cat("CV AUC:", round(clf$performance$auc, 3), "\n")
head(clf$feature_importance)
```

---

# Visualization

```{r plots, eval=FALSE}
# UMAP colored by disease state
plotCoMorTrax(cmt, type = "umap", colorBy = "disease_state")

# Interaction heatmap of top genes
plotCoMorTrax(cmt, type = "interactionHeatmap", topN = 50)

# Volcano: interaction delta vs -log10(p)
plotCoMorTrax(cmt, type = "volcanoInteraction", labelTopN = 15)

# Cell vulnerability ranking
plotCoMorTrax(cmt, type = "vulnerabilityRank")

# Classifier ROC curve
plotCoMorTrax(cmt, type = "classifierROC")

# Full overview (all key plots)
plotCoMorTrax(cmt, type = "overview")
```

---

# Validation

```{r validate}
# Known disease genes from literature
known <- c("APOE", "CLU", "TREM2", "BIN1",   # AD GWAS hits
           "TCF7L2", "PPARG", "KCNJ11")        # T2D GWAS hits

val <- validateCoMorTrax(
  cmt,
  knownGenes   = known,
  nBootstrap   = 100,
  verbose      = TRUE
)

cat("Hypergeometric p-value:", val$biological$pvalue, "\n")
cat("Bootstrap stability:   ", round(val$robustness, 3), "\n")
```

---

# Exporting Results

```{r export}
exportCoMorTrax(
  cmt,
  outDir  = "./CoMorTrax_results",
  formats = c("csv", "rds"),
  verbose = TRUE
)
```

Output directory structure:

```
CoMorTrax_results/
├── CoMorTraxObject.rds
├── disease_matrix.csv
├── interaction_scores.csv
├── de_results/
│   └── <comparison>_<cellType>.csv
├── pathway_enrichment/
│   └── <geneSet>_<database>.csv
├── cell_vulnerability.csv
└── network_rewiring.csv
```

---

# Complete Minimal Example

```{r full-example}
library(CoMorTrax)
set.seed(2026)

# Simulate
sce <- simulateCoMorTrax(n_cells = 200, n_genes = 150,
                          n_syn = 10, n_ant = 8,
                          diseases = c("AD", "T2D"))

# Full pipeline in one chain
cmt <- createCoMorTraxObject(sce)                           |>
  encodeDiseaseLabels(diseaseCol = c("disease_AD",
                                      "disease_T2D"))       |>
  runComorbidityDE(method = "wilcoxon")                     |>
  quantifyInteractionEffects(nPermutations = 200)           |>
  scoreCellVulnerability(includePathwayScore = FALSE)       |>
  buildComorbidityNetworks(maxGenes = 100)                  |>
  trainComorbidityClassifier(method = "randomForest",
                              cvFolds = 3L)

# Summary
summaryCoMorTrax(cmt)
```

---

# Session Information

```{r session}
sessionInfo()
```

---

# References

- Amezquita RA et al. (2020). Orchestrating single-cell analysis with
  Bioconductor. *Nature Methods*, 17, 137-145.

- McCarthy DJ et al. (2017). Scater: pre-processing, quality control,
  normalisation and visualisation of single-cell RNA-seq data in R.
  *Bioinformatics*, 33, 1179-1186.

- Robinson MD et al. (2010). edgeR: a Bioconductor package for differential
  expression analysis of digital gene expression data. *Bioinformatics*,
  26, 139-140.

- Salem A (2026). CoMorTrax: A Comorbidity-Aware Multi-Scale Transcriptomic
  Analysis Framework. Bioconductor package version 0.99.0.
