---
title: "CoMorTrax: Comorbidity-Aware Multi-Omics Disease Modeling at Single-Cell Resolution"
author:
  - name: Ahmed Mokhtar Ramzy Salem
    affiliation: Bioinformatics Research Group
    email: ahmed.salem@bioinformatics.org
date: "`r Sys.Date()`"
package: CoMorTrax
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{CoMorTrax Workflow: Comorbidity-Aware Disease Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  eval = FALSE
)
```

# Introduction

## The Challenge of Comorbidity in Single-Cell Studies

Comorbidities — the co-occurrence of two or more diseases in the same patient — are
extraordinarily common in clinical medicine, yet represent one of the most underexplored
areas in single-cell genomics. Existing computational frameworks analyze each disease
as an independent entity, comparing "Disease A vs. Control" or "Disease B vs. Control."
This reductionist approach fundamentally fails to capture the molecular crosstalk that
occurs when diseases co-exist within the same cellular environment.

Consider a patient with both Alzheimer's disease (AD) and Type 2 Diabetes (T2D). The
transcriptomic signature of neurons or microglia in this patient is not simply the sum
of the AD signature and the T2D signature. Gene regulatory networks rewire; pathways
that are synergistically activated under comorbidity may not be detectable in either
single-disease state. This is the molecular basis of why comorbid patients often have
worse outcomes than predicted by their individual conditions alone.

## What CoMorTrax Provides

**CoMorTrax** (*Comorbidity Tracking*) is the first dedicated Bioconductor framework
for explicitly modeling these multi-disease interactions at single-cell resolution.

### Key Innovations

| Feature | Conventional Tools | CoMorTrax |
|---|---|---|
| Disease modeling | Single disease at a time | Multi-label comorbidity encoding |
| DE analysis | Disease vs. Control | All pairwise + comorbid comparisons |
| Interaction effects | Not captured | Explicit synergy/antagonism scoring |
| Pathway analysis | Standard enrichment | Comorbidity-stratified enrichment |
| Cell vulnerability | Not quantified | Composite vulnerability scoring |
| Network analysis | Static GRNs | Rewiring between disease states |
| ML prediction | Single disease | Multi-label comorbidity prediction |

## Installation

CoMorTrax is available from Bioconductor:

```{r install-bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("CoMorTrax")
```

For the development version from GitHub:

```{r install-github, eval=FALSE}
BiocManager::install("AhmedSalem/CoMorTrax")
```

## Loading the Package

```{r load}
library(CoMorTrax)
library(SingleCellExperiment)
library(ggplot2)
```

---

# Quick Start

## Simulating Example Data

CoMorTrax ships with a simulation engine that generates realistic comorbidity data
with injected synergistic/antagonistic signals for benchmarking and exploration:

```{r simulate}
set.seed(42)
sce <- simulateCoMorTrax(
  n_cells    = 1000,
  n_genes    = 500,
  n_syn      = 30,     # synergistic genes (non-additive amplification)
  n_ant      = 20,     # antagonistic genes (non-additive suppression)
  cell_types = c("Neuron", "Microglia", "Astrocyte", "OPC"),
  diseases   = c("AD", "T2D", "Dementia"),
  seed       = 42
)

sce
```

```{r show-sim, echo=FALSE}
# Expected output:
# class: SingleCellExperiment
# dim: 500 1000
# metadata(1): params
# ...
# colData names(8): cell_type disease_AD disease_T2D disease_Dementia ...
```

---

# Full Workflow

## Step 1 — Create a CoMorTrax Object

### From an Existing SingleCellExperiment

If you have already preprocessed data:

```{r create-from-sce}
cmt <- createCoMorTraxObject(
  data   = sce,
  params = list(
    minCells   = 3,
    minGenes   = 200,
    maxMitoPCT = 20
  )
)
cmt
```

### From a Raw Count Matrix

```{r create-from-matrix, eval=FALSE}
# Assuming 'counts_mat' is a genes × cells matrix
# and 'cell_meta' is a DataFrame with disease columns

cmt <- createCoMorTraxObject(
  data      = counts_mat,
  colData   = cell_meta,
  rowData   = gene_meta
)
```

## Step 2 — Quality Control and Preprocessing

The `preprocessCoMorTrax()` function performs a complete, reproducible preprocessing
pipeline including QC metric computation, mitochondrial gene filtering, normalization,
highly variable gene selection, PCA, optional batch correction, and UMAP embedding.

```{r preprocess}
cmt <- preprocessCoMorTrax(
  object         = cmt,
  normMethod     = "logNorm",    # "logNorm", "scran", or "scTransform"
  nHVG           = 2000,
  nPCs           = 30,
  batchCorrect   = FALSE,        # Set TRUE if 'batch' column present in colData
  batchMethod    = "harmony",    # "harmony" or "fastMNN"
  runUMAP        = TRUE,
  verbose        = TRUE
)
```

## Step 3 — Disease Label Encoding

This is the pivotal step that distinguishes CoMorTrax from conventional workflows.
We map each cell to its complete disease profile, creating a binary disease matrix.

### Encoding from Multiple Binary Columns

If your `colData` contains separate binary columns per disease (e.g., `disease_AD`,
`disease_T2D`):

```{r encode-binary}
cmt <- encodeDiseaseLabels(
  object       = cmt,
  diseaseCol   = c("disease_AD", "disease_T2D", "disease_Dementia"),
  controlLabel = "control",
  verbose      = TRUE
)
```

### Encoding from a Single Categorical Column

```{r encode-categorical, eval=FALSE}
# If colData has a 'diagnosis' column with values like "AD;T2D", "Control", etc.
cmt <- encodeDiseaseLabels(
  object       = cmt,
  diseaseCol   = "diagnosis",
  separator    = ";",
  controlLabel = "Control"
)
```

### Inspecting the Disease Matrix

```{r inspect-disease}
# View the binary disease matrix (cells × diseases)
head(diseaseMatrix(cmt))

# View all comparison groups defined
str(diseaseGroups(cmt))

# Enumerate all pairwise + comorbid comparisons
listComorbidityComparisons(cmt)
```

Typical output from `listComorbidityComparisons()`:

```
## Comorbidity Comparisons Available:
## ──────────────────────────────────────────────────────────
## Single-disease comparisons (vs. Control):
##   [1] AD_vs_Control
##   [2] T2D_vs_Control
##   [3] Dementia_vs_Control
## Comorbid comparisons (vs. Control):
##   [4] AD_T2D_vs_Control
##   [5] AD_Dementia_vs_Control
##   [6] T2D_Dementia_vs_Control
##   [7] AD_T2D_Dementia_vs_Control
## Comorbid vs. single-disease:
##   [8] AD_T2D_vs_AD
##   [9] AD_T2D_vs_T2D
##  [10] AD_Dementia_vs_AD
## ...
## Total: 15 comparisons enumerated
```

## Step 4 — Comorbidity-Aware Differential Expression

Run differential expression across all comparisons, stratified by cell type:

```{r run-de}
cmt <- runComorbidityDE(
  object      = cmt,
  cellTypeCol = "cell_type",
  comparisons = "auto",      # Run all enumerated comparisons
  method      = "wilcoxon",  # "wilcoxon", "edgeR", or "MAST"
  minCells    = 10,
  logFCcutoff = 0.5,
  adjPcutoff  = 0.05,
  verbose     = TRUE
)
```

### Accessing DE Results

```{r access-de}
# Get DE table for a specific comparison and cell type
de_table <- getDeTable(
  cmt,
  comparison = "AD_T2D_vs_Control",
  cellType   = "Neuron"
)

head(de_table)
#     gene     logFC    pvalue      padj significant direction
# 1  APOE4  2.341    1.2e-08  3.4e-06        TRUE          up
# 2  CLU    1.892    4.5e-07  8.9e-05        TRUE          up
# 3  TREM2 -1.203    2.3e-06  3.1e-04        TRUE        down
# ...
```

## Step 5 — Interaction Effect Quantification ★ Core Innovation

This is the heart of CoMorTrax. For each gene, we compute the **interaction delta**:

$$\Delta_{interaction} = \text{FC}(D_1 \cap D_2 \text{ vs. Ctrl}) - [\text{FC}(D_1 \text{ vs. Ctrl}) + \text{FC}(D_2 \text{ vs. Ctrl})]$$

A positive $\Delta$ indicates **synergistic** amplification under comorbidity.
A negative $\Delta$ indicates **antagonistic** suppression.
A $|\Delta|$ near zero indicates **additive** (independent) effects.

```{r interaction-effects}
cmt <- quantifyInteractionEffects(
  object         = cmt,
  disease1       = "AD",
  disease2       = "T2D",
  cellTypeCol    = "cell_type",
  synThreshold   = 0.5,      # |Δ| > 0.5 classified synergistic
  antThreshold   = -0.5,     # Δ < -0.5 classified antagonistic
  nPermutations  = 1000,     # Permutation-based p-values
  adjPcutoff     = 0.05,
  verbose        = TRUE
)
```

### Retrieving Interaction Gene Lists

```{r interaction-genes}
# Synergistic genes (amplified in comorbidity beyond expectation)
syn_genes <- getSynergisticGenes(cmt, cellType = "Neuron")
cat("Synergistic genes in Neurons:", length(syn_genes), "\n")

# Antagonistic genes
ant_genes <- getAntagonisticGenes(cmt, cellType = "Microglia")

# Additive genes
add_genes <- getAdditiveGenes(cmt)

# Full interaction table
int_table <- interactionScores(cmt)
head(int_table[order(abs(int_table$interactionDelta), decreasing=TRUE), ])
```

## Step 6 — Pathway-Level Comorbidity Enrichment

Run enrichment analysis separately on synergistic and antagonistic gene sets.
This reveals pathways uniquely activated/suppressed **only under comorbidity**:

```{r pathway-analysis}
cmt <- runComorbidityPathways(
  object     = cmt,
  genesets   = c("GO_BP", "KEGG", "Reactome"),
  organism   = "hsa",        # "hsa" for human, "mmu" for mouse
  minGSSize  = 10,
  maxGSSize  = 500,
  pCutoff    = 0.05,
  verbose    = TRUE
)
```

### Comparing Pathway Enrichment Across Gene Sets

```{r pathway-comparison}
# Side-by-side comparison: synergistic vs. antagonistic pathways
path_comparison <- comparePathways(
  cmt,
  cellType = "Neuron",
  topN     = 20
)

head(path_comparison)
```

## Step 7 — Cell-Type Vulnerability Scoring

Score each cell type by its degree of comorbidity interaction burden:

```{r cell-vulnerability}
cmt <- scoreCellVulnerability(
  object      = cmt,
  cellTypeCol = "cell_type",
  weights     = list(
    synergistic  = 0.35,   # weight for number of synergistic genes
    antagonistic = 0.25,   # weight for number of antagonistic genes
    pathway      = 0.25,   # weight for pathway disruption score
    magnitude    = 0.15    # weight for mean |interaction delta|
  ),
  normalize   = TRUE
)

# View vulnerability rankings
vul_scores <- cellVulnerability(cmt)
vul_scores[order(vul_scores$vulnerabilityScore, decreasing=TRUE), ]
```

Typical output:
```
##   cellType  nSynGenes  nAntGenes  pathwayScore  meanDelta  vulnerabilityScore  rank
## 1  Neuron         28         18          0.89     1.23            0.94           1
## 2  Microglia      21         14          0.73     0.98            0.76           2
## 3  Astrocyte      14          9          0.51     0.71            0.54           3
## 4  OPC             6          4          0.22     0.41            0.23           4
```

## Step 8 — Gene Regulatory Network Modeling

Build comorbidity-stratified gene regulatory networks and detect rewiring events:

```{r networks}
cmt <- buildComorbidityNetworks(
  object      = cmt,
  method      = "correlation",   # "correlation", "WGCNA", or "GENIE3"
  states      = c("AD", "T2D", "AD_T2D", "Control"),
  corMethod   = "pearson",
  corCutoff   = 0.6,
  pCutoff     = 0.01,
  verbose     = TRUE
)
```

### Analyzing Network Rewiring

```{r rewiring}
# Compare network topology between two states
rewiring <- computeRewiring(
  cmt,
  state1   = "Control",
  state2   = "AD_T2D",
  topHubs  = 20
)

# Get hub genes
hubs <- getNetworkHubs(cmt, state = "AD_T2D", top = 20)
print(hubs)
```

## Step 9 — Machine Learning Comorbidity Classifier

Train a classifier to predict comorbidity state from gene expression:

```{r ml-classifier}
cmt <- trainComorbidityClassifier(
  object          = cmt,
  method          = "randomForest",   # "randomForest", "lasso", or "elasticNet"
  featureSource   = "interaction",    # use interaction genes as features
  nTrees          = 500,
  kFold           = 5,
  verbose         = TRUE
)

# View classifier performance
clf_results <- classifier(cmt)
cat("Cross-validated AUC:", round(clf_results$cv_auc, 3), "\n")
cat("Balanced Accuracy:", round(clf_results$balanced_accuracy, 3), "\n")
```

## Step 10 — Spatial Mapping (Optional)

If you have spatial transcriptomics data, map comorbidity interaction signals
onto tissue coordinates:

```{r spatial, eval=FALSE}
cmt <- mapSpatialComorbidity(
  object       = cmt,
  spatialData  = spatial_sce,   # SpatialExperiment object
  method       = "mean_expr",   # "mean_expr" or "ssgsea"
  geneSet      = "synergistic"
)

# Visualize on tissue
plotSpatialComorbidity(cmt, title = "AD+T2D Synergistic Score")
```

---

# Visualization

CoMorTrax provides a unified `plotCoMorTrax()` interface for all visualizations:

## UMAP with Disease Combinations

```{r plot-umap}
plotCoMorTrax(cmt, type = "umap",
  colorBy = "disease_state",
  title   = "Cells Colored by Comorbidity State"
)
```

## Interaction Effect Heatmap

```{r plot-heatmap}
plotCoMorTrax(cmt, type = "interactionHeatmap",
  cellType = "Neuron",
  topN     = 50,
  title    = "Top 50 Interaction Genes — Neurons"
)
```

## Volcano Plot of Interaction Scores

```{r plot-volcano}
plotCoMorTrax(cmt, type = "volcanoInteraction",
  cellType     = "Neuron",
  labelTopN    = 15,
  synColor     = "#E63946",
  antColor     = "#457B9D",
  addColor     = "#6c757d"
)
```

## Pathway Dot Plot

```{r plot-pathways}
plotCoMorTrax(cmt, type = "pathwayDotplot",
  cellType  = "Neuron",
  geneSet   = "synergistic",
  topN      = 15
)
```

## Cell Vulnerability Ranking

```{r plot-vulnerability}
plotCoMorTrax(cmt, type = "vulnerabilityRank",
  fillColor = "#E63946",
  title     = "Cell-Type Vulnerability to AD+T2D Comorbidity"
)
```

## Classifier ROC Curve

```{r plot-roc}
plotCoMorTrax(cmt, type = "classifierROC",
  title = "Comorbidity Classifier Performance"
)
```

## Full Summary Overview

```{r plot-overview}
plotCoMorTrax(cmt, type = "overview")
```

---

# Validation

CoMorTrax includes tools for multi-layer biological and technical validation:

```{r validate}
val_results <- validateCoMorTrax(
  object       = cmt,
  knownGenes   = c("APOE", "CLU", "TREM2", "BIN1",  # AD genes
                   "TCF7L2", "PPARG", "KCNJ11"),      # T2D genes
  nBootstrap   = 100,
  verbose      = TRUE
)

# Print validation summary
print(val_results)
```

---

# Exporting Results

```{r export}
exportCoMorTrax(
  object    = cmt,
  outDir    = "./CoMorTrax_results",
  format    = c("csv", "rds"),
  compress  = TRUE
)
```

This creates a structured output directory:

```
CoMorTrax_results/
├── interaction_effects/
│   ├── interaction_table.csv
│   ├── synergistic_genes.csv
│   └── antagonistic_genes.csv
├── differential_expression/
│   ├── AD_T2D_vs_Control_Neuron.csv
│   └── ...
├── pathway_enrichment/
│   ├── synergistic_GO_BP.csv
│   └── antagonistic_KEGG.csv
├── cell_vulnerability/
│   └── vulnerability_scores.csv
├── networks/
│   └── network_rewiring.csv
├── classifier/
│   └── classifier_performance.csv
└── CoMorTrax_object.rds
```

---

# Complete End-to-End Example

```{r full-example}
library(CoMorTrax)
set.seed(2024)

# 1. Simulate data with known comorbidity signal
sce <- simulateCoMorTrax(
  n_cells = 800, n_genes = 400,
  n_syn = 25, n_ant = 15,
  diseases = c("AD", "T2D")
)

# 2–3. Create object + preprocess
cmt <- createCoMorTraxObject(sce) |>
  preprocessCoMorTrax(normMethod = "logNorm", nHVG = 1500)

# 4. Encode disease labels
cmt <- encodeDiseaseLabels(cmt, diseaseCol = c("disease_AD","disease_T2D"))

# 5. Comorbidity-aware DE
cmt <- runComorbidityDE(cmt, method = "wilcoxon", minCells = 5)

# 6. Interaction effects
cmt <- quantifyInteractionEffects(
  cmt, disease1 = "AD", disease2 = "T2D",
  nPermutations = 500
)

# 7. Pathway enrichment
cmt <- runComorbidityPathways(cmt, genesets = "GO_BP")

# 8. Vulnerability scoring
cmt <- scoreCellVulnerability(cmt)

# 9. Network modeling
cmt <- buildComorbidityNetworks(cmt, method = "correlation")

# 10. ML classifier
cmt <- trainComorbidityClassifier(cmt, method = "randomForest")

# Full summary
summaryCoMorTrax(cmt)

# Export
exportCoMorTrax(cmt, outDir = "./my_results")
```

---

# Session Information

```{r session-info}
sessionInfo()
```

---

# References

<!-- Bibliography is auto-generated from references.bib -->

Amezquita, R.A., et al. (2020). Orchestrating single-cell analysis with Bioconductor.
*Nature Methods*, 17, 137–145.

Hao, Y., et al. (2021). Integrated analysis of multimodal single-cell data. *Cell*, 184, 3573–3587.

Love, M.I., et al. (2014). Moderated estimation of fold change and dispersion for RNA-seq
data with DESeq2. *Genome Biology*, 15, 550.

McCarthy, D.J., et al. (2017). Scater: pre-processing, quality control, normalisation
and visualisation of single-cell RNA-seq data in R. *Bioinformatics*, 33, 1179–1186.

Salem, A.M.R. (2024). CoMorTrax: A comorbidity-aware multi-omics framework for modeling
disease interactions at single-cell resolution. *bioRxiv*.
